Remotely sensing river greenhouse gas exchange velocity using the SWOT satellite

### 2022-03-09 docs/manuscript_v9.Rmd
manuscript v10

## Abstract
Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gas via gas exchange. Despite much progress, there is still large uncertainty in the temporal dynamics of gas exchange and thus carbon emissions to the atmosphere. Much of this uncertainty stems from uncertainty in gas exchange velocity $k_{600}$ (the rate of this diffusive transport). We propose that the NASA/CNES/USKA/CSA SWOT satellite, set to usher in a new age of hydrology research at the global scale, can provide similarly transformative insights to fluvial gas exchange modeling upon launch in 2022. Here, we present work inferring $k_{600}$ from synthetic future SWOT observations without in situ calibration. We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$ We then couple this model with established Bayesian inference techniques to develop BIKER, the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 166 discrete measurements of river gas exchange, yielding an algorithm that predicts the $k_{600}$ timeseries solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`. Similar to Bayesian remote sensing algorithms for SWOT, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than the absolute rates of exchange (median relative bias of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$MAE), 2)*100`%). Finally, we show BIKER is robust to measurement errors implicit in the SWOT data. With SWOT set to launch in late 2022, we suggest that BIKER will be useful in mapping the global-scale spatiotemporal dynamics of fluvial gas exchange in large rivers.

@@

There is a robust existing literature exploring spatiotemporal patterns in $\delta CO_2_{water}$ [e.g. @ahoDistinctConcentrationdischargeDynamics2021; @ahoDifferentialResponseGreenhouse2019; @crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000; @liuHydrologicControlsPCO22018a]. This work has lead to recent river-reach explicit modeling of $\delta CO_2$ using global hydrography datasets at up to monthly temporal resolutions [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; *Liu accepetd*], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking. As equations 1 and 2 dictate, $\delta CO_2_{water}$ and $k_{CO_2}$ share the burden of calculating $FCO_2$ and therefore these next-generation models will have to contend with this discrepancy, which is directly dictated by the lack of direct measurements of global river hydraulics.

@@

Applying these *k* models across watersheds, regions, or continents is called ‘upscaling.’ This upscaling gives the impossible-to-measure *k* term in Equation 1 for any arbitrary number of rivers, but also changes the base parameters that ultimately control the final estimate of $FCO_2$. That is, by making *k* a function of hydraulics, $FCO_2_{upscaled}$ is now a direct function of river hydraulics. This functional relationship is described in equation 2. It suggests that $FCO_2$ estimates are not only at the mercy of the accuracy and spatiotemporal resolution of $\delta CO_2$, but also the accuracy and resolution of our river hydraulics estimates.

@@

Global upscalng has been attempted inthe literature. @raymondGlobalCarbonDioxide2013, @lauerwaldSpatialPatternsCO22015, and @horgbyUnexpectedLargeEvasion2019 all relied on $k_{CO_2}$ values indirectly estimated using mean annual streamflow models and hydraulic scaling equations to predict the hydraulic terms to in turn predict $k_{CO_2}$, while @borgesGloballySignificantGreenhousegas2015 used a combination of the above method and a constant $k_{CO_2}$ in space and time to upscale over Africa. In all of these foundational studies, the temporal dynamics of $k_{CO_2}$ (and thus dynamics in $FCO_2$) were ignored because of hydraulic data limitations. It has been shown at the field-scale that temporal dynamics of gas exchange can vary widely from site to site [@wallinSpatiotemporalVariabilityGas2011], but it has remained impractical to obtain temporally explicit $k_{CO_2}$ at continental-to-global scales. More recently, *Liu etal accepted* performed a first assessment of monthly temporal dynamics in the global river $CO_2$ flux, though they relied on modeled streamflow and used the same model for $k_{600}$ as previous studies [@raymondGlobalCarbonDioxide2013] to achieve this.

@wangPhysicallyBasedScaling2021 recently attempted to address this global *k* problem by simulating $k_{O_2}$ in 35 rivers of many sizes (widths ranging from 0.23–349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to infer what $K_{O_2}$ must have been to produce their 'observations'. They then compared this simulated dataset against direct measurements of *k*, finding similar performance and parameter values against process-based models of gas evasion. However, they were still limited by a lack of direct hydraulic measurements and had to rely on hydraulic scaling equations to estimate river depth and velocity. Even though approaches like @wangPhysicallyBasedScaling2021's are incredibly useful for expanding our mechanistic understanding of gas exchange, they are less useful for global upscaling purposes as they rely on highly detailed in situ DO data for every river [@hallGasExchangeStreams2020].

We have established that literature has a reasonably good understanding of $\delta CO_2$ and a poor understanding in $k_{600}$ (and therefore $k_{CO_2}$) across large areas and in time. In theory, the discrepancy between the quality of our $\delta CO_2_{water}$ and $k_{CO_2}$ estimates could be alleviated if direct hydraulics measurements (and in turn $k_{CO_2}$ via equation 2) were available at the global scale at a sufficient temporal resolution. This is turn would also address the uncertainty regarding continental-to-global scale temporal dynamics of gas exchange noted earlier. Conveniently, these hydraulic data will soon be available via the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission. SWOT is expected to launch in 2022 and provide the world's first direct measurements of global water surface extent and elevation (and therefore water surface slope) at novel temporal resolutions. SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle, measuring rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. Via its direct hydraulic measurements, SWOT is expected to usher in a sea change in global-scale hydrology, and could similarly transform fluvial biogeochemsitry if techniques are developed to ingest SWOT data and infer *k*. In this context, we borrow tools from fluvial geomorphology and existing SWOT algorithms to answer the following two questions:

@@

To answer this first question, we use hydraulic geometry- the fundamental geomorphic relationships between streamflow and channel shape [@leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] to develop a process-based model for large-river (here defined as wider than 50m to align with SWOT) $k_{600}$. We then take these findings and explore the second question by implementing this hydraulic model, which in turn defines $k_{600}$, within an algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate') to infer $k_{600}$ solely from SWOT measurements. The goal of BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and would improve results) such that it is globally implementable on any SWOT-observable river. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (as SWOT has not yet launched) and explore BIKER's robustness to the expected measurement errors implicit in the satellite's observations. Finally, we also couple BIKER's $k_{600}$ estimates with $\delta CO_2$ to predict gas fluxes and compare these against established literature methods that are reliant on in situ hydraulic measurements.

This paper is split into two distinct parts: gas-exchange theory/model development (section 2) and BIKER algorithm development/validation (section 3). Section 3 is fundamentally dependent on the results presented in section 2, therefore section 2 presents both theory and results. Both sections detail the data used. We conclude with a discussion (section 4) that encompasses both gas exchange theory and remote sensing. We also provide a flowchart detailing the entire study as Figure 1.

![Figure 1: Flowchart detailing the BIKER algorithm. Dashed lines indicate workflow to calculate the observed data that we validate against. See Appendix A for variable nomenclature used in this study.](cache/flowchart_2.jpg)

@@

### 2.3 Deriving a large-river $k_{600}$ model
Given the theoretical context provided in Section 2.2, we now turn to SWOT-observable systems specifically. Rivers and streams change predictably along their longitudinal profile from headwater to ocean, and we can exploit the hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems. In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e. $R_h \approx H$ **[CITE]**. This is a common assumption in hydraulic and geomorphic modeling of large rivers, and flows in a SWOT-observable river flow have an average $\frac{R_h}{H}$ ratio of `r HG_swot$mean` and standard deviation of `r HG_swot$sd` (n = `r HG_swot$n`; see Text S1 for how we built this dataset). We refer to these rivers as 'hydraulically-wide'.

We therefore assume that all SWOT-observable rivers are hydraulically-wide to derive a model for gas exchange. The overall goal is to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT. To do this, we impose $\epsilon=\epsilon_D$ and $R_H=H$ on the Reynolds extension model and arrive at equation 7 (with statistical coefficient $\beta_1$). Equation 7 thus define gas exchange velocity solely as a function of slope, mean flow depth, and mean flow velocity. This is theoretically valid only in a hydraulically-wide channel.

@@

We also test the performance of three other models for predicting *k* in hydraulically-wide channels via the other three unique combinations of equations 3-4 and equations 5-6. T While the complete model derivations and results for the four models are provided in Text S2 and Figure S1, the final and best-performing model (equation 7) is presented and used below.

### 2.4 Model validation
With equation 7 derived, we now empirically its strength of fit for hydraulically-wide river flows. We validate on the dataset of in situ measurements of $k_{600}$, after filtering for measurements made in hydraulically-wide channels, which was defined as flows whose hydraulic radius was within 1% of their mean flow depth. All told, this amounts to `r n` direct measurements of hydraulically-wide $k_{600}$. Equation 7 is assessed via the coefficient of determination ($r^2$) and plotted in Figure 2. Note that Figure 2 axes are plotted in logarithmic space only for visualization: model fit and validation (via $r^2$) were calculated in linear space as their models dictate.

@@

## 3 Exploiting equation 7 to remotely sense gas exchange velocity
We have shown that scaling $k_{600}$ via equation 7 should provide a reliable basis for hydraulically-wide rivers with a defined error structure. Further, equation 7 has only three non-directly-remotely-sensible terms: $k_{600}$, mean flow depth, and mean flow velocity. Conveniently, techniques to simultaneously infer mean flow depth and velocity from SWOT data (among other parameters) have been established over the last decade to infer streamflow from SWOT's measurements [e.g. **GLEASON & SMITH 2014**; @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018]. For BIKER, we follow the work developed by @hagemannBAMBayesianAMHGManning2017, @brinkerhoffConstrainingRemoteRiver2020, and @durandEstimatingReachaveragedDischarge2014 to infer $k_{600}$, channel depth and velocity from SWOT observations using a modified form of Manning's equation. Following a description of the data used (section 3.1), we detail algorithm development and experimental design (section 3.2) and then we present the validation results (section 3.3).

### 3.1 Data
![Figure 3: Map of the 47 hydraulic models and 1 timeseries of $CO_2$ samples used in this study. Note that hydraulic model locations are approximate as some of the models are not geo-referenced.](cache/maps/map3.jpg)

To validate BIKER, we cannot use actual SWOT measurements as SWOT has yet to launch. In the hydrology literature, it has become standard practice to benchmark SWOT-related algorithms on "SWOT-like" data [@durandIntercomparisonRemoteSensing2016]. We use 47 SWOT-simulated rivers for validation, where these simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights, slopes, and widths are labelled as RS observations and are used as the sole inputs to BIKER. These datasets are created using standard hydraulic models forced with known inflows and measured bathymetry to model the hydraulic response of the rivers, and then those terms visible to SWOT were extracted to produce hydraulically realistic synthetic observations. These data were published by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016 as benchmarking datasets to explore remote sensing of discharge (RSQ) algorithm performance for the SWOT mission.

@@

For $CO_2$ evasion and carbon emissions calculations, we use 26 bi-weekly dissolved $CO_2$ samples collected by @beaulieuControlsGasTransfer2012 at one location in the Ohio River for one calender year from 2008-2009 (Figure S2). Note that this $CO_2$ data is for the Ohio River only but was applied to all rivers to provide a physically realistic signal for $CO_2$ fluxes with meaningful seasonality and dynamics. Therefore, the raw carbon emissions estimates presented in this paper are completely meaningless in the context of actually measured carbon emissions from these rivers, but are better than specifying $CO_2$ concentrations devoid of context. These data are necessary as we are interested in the effect of BIKER $k_{600}$ errors on eventual fluxes and comparing these fluxes with published methods. Therefore, applying these 'unit' $CO_2$ values allows for such a comparison by providing a realistic timeseries.

@@

$$p(\Theta|x) = \frac{f(x|\Theta)p(\Theta)}{p(x)} \mathbf{(8)}$$
The heart of BIKER is its reformulation of the $k_{600}$ model (equation 7) as a Bayesian sampling model that is conditional on the non-remotely-sensible parameters (i.e. $f(x|\Theta)$). This is similar to the 'McFLI' ('Mass-Conserved Flow Law Inversion') logic used in some SWOT RSQ algorithms [@gleasonTrackingRiverFlows2017]. To start, we write $k_{600}$ as a function of SWOT-observables $W$ and $S$. This algebra is carried out using equation 7, the fitted value for $\beta_1$ from Figure 2 (`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$coef,2)`), and Manning's equation for mean flow velocity ($\bar{U}=\frac{1}{n}R_h^{2/3}S^{1/2}$). Following section 2.3, we continue to assume that the channel is hydraulically-wide ($R_h = H =\frac{A}{W}$). To leverage additional SWOT data, the wetted channel area *A* is further split into the the SWOT-unobservable portion $A_0$ and SWOT-observable portion $dA$ following @durandEstimatingReachaveragedDischarge2014 and @hagemannBAMBayesianAMHGManning2017 where $dA_{it} = \sum_{t':W_{t'}\le W_t}{W_{t'} \delta H_{t'}}$ for cross-section *i* and timestep *t*.

@@

$$\left[log(62.82) + \frac{7}{16}log(g) + \frac{9}{16}log(S_{i,t}) - \frac{35}{48}log(W_{i,t})\right] \sim normal\left(\left[log(k_{600_t})+\frac{1}{4}log(n_i)-\frac{35}{48}log(A_0_i+dA_{i,t})\right], \sigma^2_{k_{600}}\right) \mathbf{(10)}$$
Equations 8 and 10 also necessitate that we specify prior distributions for the parameters $A_0_i$, $k_{600_t}$, and $n_i$. Prior distributions formalize the a priori estimates and uncertainties for the non-remotely-sensed terms. More intuitively, BIKER priors represent our 'prior river knowledge' of what $A_0_i$, $k_{600_t}$, and $n_i$ probably are for some river since they cannot be directly remotely sensed. This is analogous to the 'empirical Bayes approach' to Bayesian inference [@hoffFirstCourseBayesian2009]. Our goal in prior specification was to rely on absolutely no in situ information such that we could run this method on any river on Earth solely using SWOT observations. In theory, more informed priors via various a priori information about a specific river will improve BIKER performance, but here we chose to test the fully generalized algorithm. Therefore, the validation presented here is a 'worst-case scenario', wherein BIKER performance should improve with better prior information on the river. In that context, we used a variation of the prior specification method developed by @brinkerhoffConstrainingRemoteRiver2020, who developed geomorphic 'river types' with distinct prior sets for $A_0_i$ and $n_i$. These priors are assigned to a river solely using SWOT data *W* and *S*, therefore meeting our needs for complete global implementability. Prior assignment for $k_{600_t}$ was developed similarly (all prior specifications are elaborated in Text S3).

@@

#### 3.2.2 BIKER validation
We validate BIKER, assuming no measurement error, on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers using daily simulated hydraulics. We also re-validate BIKER under the 'measurement-error' scenario using the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers with [@frassonExploringFactorsControlling2021]'s SWOT error model to corrupt the hydraulics to mimic realistic SWOT measurements (widths, heights, and slopes). These data were outlined in section 3.1.

>>>>

## Abstract
Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gas via gas exchange. Despite much progress, there is still large uncertainty in the temporal dynamics of gas exchange and thus carbon emissions to the atmosphere. Much of this uncertainty stems from uncertainty in gas exchange velocity $k_{600}$ (the rate of this diffusive transport). We propose that the NASA/CNES/USKA/CSA SWOT satellite can provide new insights to fluvial gas exchange modeling upon launch in 2022. Here, we present work inferring $k_{600}$ from synthetic future SWOT observations without in situ calibration. We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$ We then couple this model with established techniques to develop BIKER, the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 166 measurements of river gas exchange, yielding an algorithm that predicts the $k_{600}$ timeseries solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`. Similar to remote sensing algorithms for SWOT, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than the absolute rates of exchange (median normalized RMSE of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$NRMSE), 2)*100`%). Finally, BIKER is robust to measurement errors implicit in the SWOT data. Upon SWOT's launch, we suggest that BIKER will be useful in mapping the global-scale spatiotemporal dynamics of fluvial gas exchange in large rivers.

@@

There is a robust existing literature exploring spatiotemporal patterns in $\delta CO_2_{water}$ [e.g. @ahoDistinctConcentrationdischargeDynamics2021; @ahoDifferentialResponseGreenhouse2019; @crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000; @liuHydrologicControlsPCO22018a]. This work has lead to recent river-reach explicit modeling of $\delta CO_2$ using global hydrography datasets at up to monthly temporal resolutions [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; *Liu accepetd*], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking.

@@

Applying these *k* models across watersheds, regions, or continents is called ‘upscaling.’ This upscaling allows for estimating the impossible-to-measure *k* term in Equation 1 for any arbitrary number of rivers, but also changes the base parameters that ultimately control the final estimate of $FCO_2$. That is, by making *k* a function of hydraulics, $FCO_2_{upscaled}$ is now a direct function of river hydraulics. This functional relationship is described in equation 2. It suggests that $FCO_2$ estimates are not only at the mercy of the accuracy and spatiotemporal resolution of $\delta CO_2$, but also the accuracy and resolution of our river hydraulics estimates.

@@

Global upscalng has been performed in the literature. @raymondGlobalCarbonDioxide2013, @lauerwaldSpatialPatternsCO22015, and @horgbyUnexpectedLargeEvasion2019 all relied on $k_{CO_2}$ values indirectly estimated using mean annual streamflow models and hydraulic scaling equations to predict the hydraulic terms to in turn predict $k_{CO_2}$, while @borgesGloballySignificantGreenhousegas2015 used a combination of the above method and a constant $k_{CO_2}$ in space and time to upscale over Africa. In all of these foundational studies, the temporal dynamics of $k_{CO_2}$ (and thus dynamics in $FCO_2$) were ignored because of hydraulic data limitations. It has been shown at the field-scale that temporal dynamics of gas exchange can vary widely from site to site [@wallinSpatiotemporalVariabilityGas2011], but it has remained impractical to obtain temporally explicit $k_{CO_2}$ at continental-to-global scales. More recently, *Liu etal accepted* performed a first assessment of monthly temporal dynamics in the global river $CO_2$ flux, though they relied on modeled streamflow and used the same model for $k_{600}$ as previous studies [@raymondGlobalCarbonDioxide2013] to achieve this.

@wangPhysicallyBasedScaling2021 recently attempted to address this global *k* problem by simulating $k_{O_2}$ in 35 rivers of many sizes (widths ranging from 0.23–349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to infer what $k_{O_2}$ must have been to produce their 'observations'. They then compared this simulated dataset against direct measurements of *k*, finding similar performance and parameter values against process-based models of gas evasion. However, they were still limited by a lack of direct hydraulic measurements and had to rely on hydraulic scaling equations to estimate river depth and velocity. Even though approaches like @wangPhysicallyBasedScaling2021's are incredibly useful for expanding our mechanistic understanding of gas exchange, they are less useful for global upscaling purposes as they rely on highly detailed in situ DO data for every river [@hallGasExchangeStreams2020].

We have established that literature has a reasonably good understanding of $\delta CO_2$ and a relatively poorer understanding of $k_{600}$ (and therefore $k_{CO_2}$) across large areas and in time. In theory, the discrepancy between the quality of our $\delta CO_2_{water}$ and $k_{CO_2}$ estimates could be alleviated if direct hydraulics measurements (and in turn $k_{CO_2}$ via equation 2) were available at the global scale at a sufficient temporal resolution. This is turn would also address the uncertainty regarding continental-to-global scale temporal dynamics of gas exchange noted earlier. Conveniently, these hydraulic data will soon be available via the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission. SWOT is expected to launch in 2022 and provide the world's first direct measurements of global water surface extent and elevation (and therefore water surface slope) at novel temporal resolutions. SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle, measuring rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. Via its direct hydraulic measurements, SWOT is expected to usher in a sea change in global-scale hydrology, and could similarly transform fluvial biogeochemsitry if techniques are developed to ingest SWOT data and infer *k*. In this context, we borrow tools from fluvial geomorphology and existing SWOT algorithms to answer the following two questions:

@@

To answer this first question, we use hydraulic geometry- the fundamental geomorphic relationships between streamflow and channel shape [@leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] to develop a process-based model for large-river (here defined as wider than 50m to align with SWOT) $k_{600}$. We then take these findings and explore the second question by implementing this hydraulic model, which defines $k_{600}$, within an algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate') to infer $k_{600}$ solely from SWOT measurements. The goal of BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and would improve results) such that it is globally implementable on any SWOT-observable river. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (as SWOT has not yet launched) and explore BIKER's robustness to the expected measurement errors implicit in the satellite's observations. Finally, we also couple BIKER's $k_{600}$ estimates with $\delta CO_2$ to predict gas fluxes and compare these against established literature methods that rely on in situ hydraulic measurements.

This paper is split into two distinct parts: gas-exchange theory/model development (section 2) and BIKER algorithm development/validation (section 3). Section 3 is fundamentally dependent on the results presented in section 2, therefore section 2 presents both theory and results. Both sections detail the data used. We conclude with a discussion (section 4) that encompasses both gas exchange theory and remote sensing. Figure 1 conceptually maps out the algorithm's approach to inferring $k_{600}$ from SWOT data.

![Figure 1: Conceptual overview of the BIKER algorithm. SWOT will directly measure water surface width and elevation (and thus slope), and the BIKER algorithm uses those observations to infer gas exchange velocity $k_{600}$ via a process-based model (section 2).](cache/flowchart3.jpg)

@@

### 2.3 Deriving a large-river $k_{600}$ model
Given the theoretical context provided in Section 2.2, we now turn to SWOT-observable systems specifically. Rivers and streams change predictably along their longitudinal profile from source to sea, and we can exploit the hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems. In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e. $R_h \approx H$ [@leopoldHydraulicGeometryStream1953]. This is a common assumption in hydraulic and geomorphic modeling of large rivers. For example, SWOT-observable river flows have an average $\frac{R_h}{H}$ ratio of `r HG_swot$mean` and standard deviation of `r HG_swot$sd` (n = `r HG_swot$n`; see Text S1 for how we built this dataset). We refer to these rivers as 'hydraulically-wide'.

We therefore assume that all SWOT-observable rivers are hydraulically-wide to derive a model for gas exchange. The overall goal is to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT. To do this, we impose $\epsilon=\epsilon_D$ and $R_H=H$ on the Reynolds extension model and arrive at equation 7 (with statistical coefficient $\beta_1$). Equation 7 thus defines gas exchange velocity solely as a function of slope, mean flow depth, and mean flow velocity. This is theoretically valid only in a hydraulically-wide channel.

@@

We also test the performance of three other models for predicting *k* in hydraulically-wide channels via the other three unique combinations of equations 3-4 and equations 5-6. While the complete model derivations and results for all four models are provided in Text S2 and Figure S1, the final and best-performing model (equation 7) is presented and used below.

### 2.4 Model validation
With equation 7 derived, we now empirically test its strength of fit in hydraulically-wide river flows. We validate on the dataset of in situ measurements of $k_{600}$, after filtering for measurements made in hydraulically-wide channels, which was defined as flows whose hydraulic radius was within 1% of their mean flow depth. All told, this amounts to `r n` direct measurements of hydraulically-wide $k_{600}$. Equation 7 is assessed via the coefficient of determination ($r^2$) and plotted in Figure 2. Note that Figure 2 axes are plotted in logarithmic space only for visualization: model fit and validation (via $r^2$) were calculated in linear space as their models dictate.

@@

## 3 Exploiting equation 7 to remotely sense gas exchange velocity
We have shown that scaling $k_{600}$ via equation 7 is useful in hydraulically-wide rivers. Further, equation 7 has only three non-directly-remotely-sensible terms: $k_{600}$, mean flow depth, and mean flow velocity. Conveniently, techniques to simultaneously infer mean flow depth and velocity from SWOT data (among other parameters) have been established over the last decade to infer streamflow from SWOT's measurements [e.g. @gleasonGlobalMappingRiver2014a; @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018]. For BIKER, we follow the work developed by @hagemannBAMBayesianAMHGManning2017, @brinkerhoffConstrainingRemoteRiver2020, and @durandEstimatingReachaveragedDischarge2014 to infer $k_{600}$, channel depth and velocity from SWOT observations using a modified form of Manning's equation. Following a description of the data used (section 3.1), we detail algorithm development and experimental design (section 3.2) and then we present the validation results (section 3.3).

### 3.1 Data
To validate BIKER, we cannot use actual SWOT measurements as SWOT has yet to launch. In the hydrology literature, it has become standard practice to benchmark SWOT-related algorithms on "SWOT-like" data [@durandIntercomparisonRemoteSensing2016]. We use 47 SWOT-simulated rivers for validation, where these simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights, slopes, and widths are labelled as RS observations and are used as the sole inputs to BIKER. These datasets are created using standard hydraulic models forced with known inflows and measured bathymetry to model the hydraulic response of the rivers, and then those terms visible to SWOT are extracted to produce hydraulically realistic synthetic observations. These data were published by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016. See Figure 3 for a map of these river's locations.

![Figure 3: Map of the 47 hydraulic models and 1 timeseries of $CO_2$ samples used in this study. Note that hydraulic model locations are approximate as some of the models are not geo-referenced.](cache/maps/map3.jpg)

@@

For $CO_2$ evasion and carbon emissions calculations, we use 26 bi-weekly dissolved $CO_2$ samples collected by @beaulieuControlsGasTransfer2012 at one location in the Ohio River (Figure 3) for one calender year from 2008-2009 (Figure S2). Note that this $CO_2$ data is for the Ohio River only but was applied to all rivers to provide a physically realistic signal for $CO_2$ fluxes with meaningful seasonality and dynamics. Therefore, the raw carbon emissions estimates presented in this paper are completely meaningless in the context of actually measured carbon emissions from these rivers, but are better than specifying $CO_2$ concentrations devoid of context. These data are necessary as we are interested in the effect of BIKER $k_{600}$ errors on eventual fluxes and comparing these fluxes with published methods. Therefore, applying these 'unit' $CO_2$ values allows for such a comparison by providing a realistic timeseries.

@@

$$p(\Theta|x) = \frac{f(x|\Theta)p(\Theta)}{p(x)} \mathbf{(8)}$$
The heart of BIKER is its reformulation of the $k_{600}$ model (equation 7) as a Bayesian sampling model that is conditional on the non-remotely-sensible parameters (i.e. $f(x|\Theta)$). This is similar to the 'McFLI' ('Mass-Conserved Flow Law Inversion') logic used in some SWOT RSQ algorithms [@gleasonTrackingRiverFlows2017]. To start, we write $k_{600}$ as a function of SWOT-observables $W$ and $S$. This algebra is carried out using equation 7, the fitted value for $\beta_1$ from Figure 2 (`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$coef,2)`), and Manning's equation for mean flow velocity ($\bar{U}=\frac{1}{n}R_h^{2/3}S^{1/2}$). Following section 2.3, we continue to assume that the channel is hydraulically-wide ($R_h = H =\frac{A}{W}$). To leverage additional SWOT data, the wetted channel area *A* is further split into the the SWOT-unobservable portion $A_0$ and SWOT-observable portion $dA$ following @durandEstimatingReachaveragedDischarge2014 and @hagemannBAMBayesianAMHGManning2017 where $dA_{it} = \sum_{t':W_{t'}\le W_t}{W_{t'} \delta H_e_{t'}}$ for cross-section *i* and timestep *t* within a mass-conserved river reach.

@@

$$\left[log(62.82) + \frac{7}{16}log(g) + \frac{9}{16}log(S_{i,t}) - \frac{35}{48}log(W_{i,t})\right] \sim normal\left(\left[log(k_{600_t})+\frac{1}{4}log(n_i)-\frac{35}{48}log(A_0_i+dA_{i,t})\right], \sigma^2_{k_{600}}\right) \mathbf{(10)}$$
Equations 8 and 10 also necessitate that we specify prior distributions for the parameters $A_0_i$, $k_{600_t}$, and $n_i$. Prior distributions formalize the a priori estimates and uncertainties for the non-remotely-sensed terms. More intuitively, BIKER priors represent our 'prior river knowledge' of what $A_0_i$, $k_{600_t}$, and $n_i$ probably are for some river since they cannot be directly remotely sensed. This is analogous to the 'empirical Bayes approach' to Bayesian inference [@hoffFirstCourseBayesian2009]. Our goal in prior specification was to rely on absolutely no in situ information such that we could run this method on any river on Earth solely using SWOT observations. In theory, more informed priors via various a priori information about a specific river will improve BIKER performance, but here we chose to test the fully generalized algorithm. Therefore, the validation presented here is a 'worst-case scenario', wherein BIKER performance should improve with better prior information on the river. In that context, we used a variation of the prior specification method developed by @brinkerhoffConstrainingRemoteRiver2020, who developed 'geomorphic river types' with distinct prior sets for $A_0_i$ and $n_i$. These priors are assigned to a river solely using SWOT data *W* and *S*, therefore meeting our needs for complete global implementability. Prior assignment for $k_{600_t}$ was developed similarly (all prior specifications are elaborated in Text S3).

@@

#### 3.2.2 BIKER validation
We validate BIKER, assuming no measurement error, on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers using daily simulated hydraulics. We also re-validate BIKER under the 'measurement-error' scenario using the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers with @frassonExploringFactorsControlling2021's SWOT error model to corrupt the hydraulics to mimic realistic SWOT measurements (widths, heights, and slopes). These data were outlined in section 3.1.
